---
title: "Editing Homicide data with Temperature Data"
author: "Nusrat Nawshin"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "show", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```
### Main Homicide Dataframe

#### reading the homicide original data

```{r}
df <- data.frame(read.csv(unz('database.zip','database.csv'), header = T))
summary(df)
```

#### Lets observe the Month column unique values

```{r}
unique(df[c("Month")])
```
#### Creating a new column named Date and mapping with Month data as (01, 02, 03 .... 11,12)

```{r, results='hide'}
require(plyr)
df$Date <- mapvalues(df$Month, 
          from=c("January","February","March", "April", "May", "June", "July", "August", "September", "October", "November", "December"), 
          to=c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12"))

summary(df)
```
#### Now lets merge year column with date and create Date column with yyyymm format to match the temperature dataframe

```{r, results='hide'}
df$Date <- paste(df$Year, df$Date, sep="")
summary(df)
```

#### Now the Date column is in yyyymm format

#### Now before merging the states lets look into the unique states in the main dataframe
```{r}
unique(df[c("State")])
```
#### There are two issues here. District of columbia is not a state so we are merging that to Maryland and the Rhode Island spelling is wrong. Renaming it from Rhodes Island to Rhode Island

```{r}
df$State_temp <- df$State
df[df$State_temp == 'District of Columbia',]$State_temp <- "Maryland"
df[df$State_temp == 'Rhodes Island',]$State_temp <- "Rhode Island"

unique(df[c("State_temp")])
```

#### Lets convert Date column back to numeric
```{r, results='hide'}
df$Date <- as.numeric(df$Date)
str(df)
```

### Joining Homicide data and Temperature data

#### Reading the temperature data

```{r}
temp_df <- data.frame(read.csv("../temperature_data/us_state_final_temp_data.csv", header = TRUE))
summary(temp_df)
```

#### Homicide data:

```{r}
summary(df)
```

#### Merging homicide data with temperature

```{r}
main_df = merge(df, temp_df, by.x=c("State_temp", "Date"), by.y=c("Location", "Date"),all.x = TRUE)
```

#### The colnames are:

```{r}
colnames(main_df)
```
#### Removing Date, State_temp and x as they ar no longer needed

### Final Homicide Data:

```{r}
main_df$State <- main_df$State_temp 
main_df <- subset(main_df, select = -c(Date, State_temp, X))
str(main_df)
```

#### Now out dataset is ready
#### Lets store it in a CSV file

```{r}
write.csv(main_df, "homicide_data.csv")
```


