---
title: "Joining all the Temperature Data part I"
author: "Nusrat Nawshin"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "show", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

### Main Dataframe

#### reading the temperature data

```{r}
df <- data.frame(read.csv("alldata1.csv", header = TRUE, skip =3))
summary(df)
```

#### only taking necessary columns: 'Location','Date','Value' -> maindf

```{r}
maindf <- subset(df, select = c(Location, Date, Value))

summary(maindf)
```

#### Renaming Value column to temp. It seems more meaningful to me.

```{r}
colnames(maindf)[which(names(maindf) == "Value")] <- "temp"
```

#### Lets look at the dataset. 
```{r}
str(maindf)
```
```{r}
unique(maindf[c("Location")])
```
#### The temperature datset dosn't have the Hawaii & Alaska states data. We have to explicitly add those states from other datasets now.

### Hawaii Data

#### Now I'm adding Hawaii datas from hawaii.xlsx

```{r}
# Loading
library("readxl")
# xlsx files
hawaii_df <- read_excel("hawaii.xlsx", skip =1)
summary(hawaii_df)
```
#### We dont need the Anual column. Lets remove it
```{r}
hawaii_df <- subset(hawaii_df, select = -c(Annual))

summary(hawaii_df)
```

#### Now lets melt the data frame

```{r}
library(reshape2)
hawaii_temp_df <- melt(hawaii_df,
                 id.vars = "Year",
                 variable.name = "Month",
                 value.name = "temp")

```


```{r}
summary(hawaii_temp_df)
```
#### Addind a new column named Location with Value "Hawaii" 
```{r}
hawaii_temp_df$Location <- "Hawaii"
str(hawaii_temp_df)
```
#### I have to convert the month column to integer so that it represents the month number
```{r}
hawaii_temp_df$Month <- as.numeric(hawaii_temp_df$Month)
```

#### Lets make each column character type
```{r}
hawaii_temp_df <- data.frame(lapply(hawaii_temp_df, as.character), stringsAsFactors=FALSE)
```

#### We have to make the month values two digits(01, 02, 03.... 11, 12) to match the other dataset format

```{r}
library(stringr)
hawaii_temp_df$Month <- str_pad(hawaii_temp_df$Month, 2, pad = "0")
str(hawaii_temp_df)
```

#### Let's join Month and year column together
```{r}
hawaii_temp_df$Date <- paste(hawaii_temp_df$Year, hawaii_temp_df$Month, sep="")
str(hawaii_temp_df)
```
#### As our desired Date structure is made. Lets remove Year and Month columns

```{r}
hawaii_temp_df <- subset(hawaii_temp_df, select = c(Location, Date, temp))
str(hawaii_temp_df)
```

#### Our Hawaii dataset has created perfectly. Now have to join Hawaii with Main data and store it in a new dataframe named us_state_temp_data

```{r}
us_state_temp_data <- rbind(maindf, hawaii_temp_df)
str(us_state_temp_data)
unique(us_state_temp_data[c("Location")])
```
#### Perfect. We have our temperature dataset with Hawaii datas. Now Lets add Alaska Data

### Alaska Data
```{r}
alaska_df <- data.frame(read.csv("alaskadata.csv", header = TRUE, skip =3))
summary(df)
str(alaska_df)
```
#### Adding a new column Location with 'Alaska' as values and rename the column 'Value' to 'temp' to match with our main temperature dataset
```{r}
alaska_df$Location <- "Alaska"
colnames(alaska_df)[which(names(alaska_df) == "Value")] <- "temp"
str(alaska_df)
```

#### Now lets join Alaska data to our all us state data

```{r}
us_state_temp_data_final <- rbind(us_state_temp_data, alaska_df)
str(us_state_temp_data_final)
unique(us_state_temp_data_final[c("Location")])
```

#### Great!! Now we have our all states temperature data

#### Lets store it in a CSV file

```{r}
write.csv(us_state_temp_data_final, "us_state_final_temp_data.csv")
```





